---
title: "Comparing metrics between vectormetrics and landscapemetrics R packages"
format: 
  html:
    toc: true
    toc_depth: 3
    number-sections: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Comparing metrics between vectormetrics and landscapemetrics R packages}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

<!--
> **Note**: This vignette uses comparison results from three data-raw scripts:
> - `11_build_metric_map.R`: Discovers metric functions in both packages
> - `12_compute_metric_comparisons.R`: Computes metrics and calculates differences
> - `13_enhance_comparison_results.R`: Adds units, categorizes differences, and provides explanations
-->

## Introduction

The **vectormetrics** package calculates landscape metrics based on vector polygon representations. 
It shares many metrics with the **landscapemetrics** package which offers landscape metrics for raster data.
In principle, both packages should produce similar or nearly identical results as long as the input data is equivalent (e.g., raster and vector representations of the exact polygon geometry).
However, differences can arise due to several factors, including computational approaches, treatment of polygon boundaries, and underlying assumptions about how landscape features are measured.

This comparison uses the `landscapemetrics::landscape` dataset, a 30×30 cell raster at 1m resolution containing 3 classes across 28 patches (9 patches in class 1, 13 in class 2, and 6 in class 3).
The same landscape was converted to vector polygons for comparison.

> **Note**: When converting polygon data to raster for analysis with **landscapemetrics**, the rasterization process affects the actual landscape structure. 
Coarser resolutions can eliminate small patches entirely, merge adjacent patches, and distort patch boundaries -- fundamentally changing what is measured.
These structural changes cascade through all dependent metrics: patch counts change, patch areas and shapes shift, which then affects perimeter-based metrics (te, ed, shape, circle, lsi), distance calculations (enn, proximity), and core area detection (cai, dcore).
This vignette uses an idealized scenario in which we have exactly the same landscape in both formats to show the pure representation difference -- but in practice, converting your polygons to raster will change the landscape structure and thus the metrics you get.

```{r visualize-datasets}
#| echo: false
#| layout-ncol: 2
library(terra)
library(sf)
library(dplyr)

# Create comparable landscapes
raster_landscape <- landscapemetrics::landscape |> terra::rast()
polygon_landscape <- raster_landscape |>
  terra::as.polygons(dissolve = TRUE) |>
  sf::st_as_sf() |>
  dplyr::rename(class = clumps) |>
  vectormetrics::get_polygon_patches("class", direction = 8) |>
  sf::st_set_crs("EPSG:2180")
# Determine number of classes
n_classes <- length(unique(polygon_landscape$class))
colors <- palette.colors(n_classes, palette = "Set2")

# Plot raster version
plot(raster_landscape, 
     main = "Raster Representation",
     col = colors,
     axes = FALSE)

# Plot vector version
plot(polygon_landscape["class"],
     main = "Vector (Polygon) Representation",
     pal = colors,
     border = "black",
     axes = FALSE)
```

```{r setup}
#| echo: false
library(vectormetrics)
library(landscapemetrics)
library(dplyr)
library(knitr)

# Load pre-computed comparison results and metric map from vignette data
vignette_data <- readRDS(system.file("extdata/vm_lsm_comparison.rds", package = "vectormetrics"))
results <- vignette_data$comparison_results_enhanced
metric_map <- vignette_data$metric_map

# Extract results by metric level
landscape_metrics <- results$landscape
class_metrics <- results$class
patch_metrics <- results$patch
all_metrics <- results$all
```

## Summary of comparison

This comparison evaluates **78 metrics** that exist in both packages, organized across three analysis levels:

- **Patch level**: `r nrow(patch_metrics |> distinct(metric_name))` metrics
- **Class level**: `r nrow(class_metrics |> distinct(metric_name))` metrics  
- **Landscape level**: `r nrow(landscape_metrics |> distinct(metric_name))` metrics

Beyond these shared metrics, **vectormetrics** offers **61 exclusive metrics** not found in **landscapemetrics** -- primarily specialized shape complexity and geometry measures suited to vector polygon analysis.

## Metrics with perfect or near-perfect agreement

Many metrics show perfect or near-perfect agreement, with absolute differences less than 0.001 on our test landscape.
These metrics are robust to the vector-raster representation difference because they rely on count-based, proportion-based, or compositional calculations that behave consistently across representations.
When patches agree at the patch level, those agreements propagate to class and landscape levels since higher levels aggregate patch-level data.

### Patch level: metrics with close agreement

```{r patch-close-agreement}
#| code-fold: true
patch_close <- patch_metrics |>
  filter(abs_diff < 0.001, class == 1) |>
  group_by(metric_name) |>
  ungroup() |>
  select(metric_name, class, n_patches_vm, n_patches_lsm, value_vm, value_lsm, abs_diff, vm_units) |>
  arrange(metric_name)

kable(
  patch_close,
  digits = 3,
  caption = sprintf("Patch-level metrics with close agreement: %d unique metrics", 
                    n_distinct(patch_close$metric_name)),
  col.names = c("Metric", "Class", "N Patches VM", "N Patches LSM", "VM Value", "LSM Value", "Abs Diff", "Units")
)
```

`r n_distinct(patch_close$metric_name)` patch-level metrics show close agreement across all patches within class 1. These foundational patch-level agreements are carried forward to higher aggregation levels.
Note that `p_core` are idential because we use the same core distance threshold -- but in practice, differences in core area detection methods (buffer vs. erosion) can lead to differences in core area metrics.

### Class level: metrics with close agreement

```{r class-close-agreement}
#| code-fold: true
class_close <- class_metrics |>
  filter(abs_diff < 0.001, class == 1) |>
  group_by(metric_name) |>
  ungroup() |>
  select(metric_name, class, value_vm, value_lsm, abs_diff, vm_units) |>
  arrange(metric_name)

kable(
  class_close,
  digits = 3,
  caption = sprintf("Class-level metrics with close agreement: %d unique metrics (class 1 shown)", 
                    n_distinct(class_close$metric_name)),
  col.names = c("Metric", "Class", "VM Value", "LSM Value", "Abs Diff", "Units")
)
```

`r n_distinct(class_close$metric_name)` class-level metrics show close agreement (class 1 shown). Class-level agreement reflects the patch-level patterns, since class metrics aggregate across patches within each class.

### Landscape level: metrics with close agreement

```{r landscape-close-agreement}
#| code-fold: true
landscape_close <- landscape_metrics |>
  filter(abs_diff < 0.001) |>
  select(metric_name, value_vm, value_lsm, abs_diff, vm_units) |>
  distinct(metric_name, .keep_all = TRUE) |>
  arrange(metric_name)

kable(
  landscape_close,
  digits = 3,
  caption = sprintf("Landscape-level metrics with close agreement (abs_diff < 0.001): %d metrics", 
                    nrow(landscape_close)),
  col.names = c("Metric", "VM Value", "LSM Value", "Abs Diff", "Units")
)
```

All `r nrow(landscape_close)` landscape-level metrics with differences below 0.001 show near-perfect agreement. Landscape-level patterns reflect the patch and class agreements that feed into them.

## Metrics with systematic differences

Some metrics differ substantially between packages due to fundamental differences in how vector and raster data represent landscape features.


### Patch level differences

```{r patch-differences}
#| code-fold: true
patch_diffs <- patch_metrics |>
  filter(abs_diff > 0.001) |>
  arrange(desc(abs_diff)) |>
  group_by(metric_name) |>
  slice(1) |>
  ungroup() |>
  select(metric_name, class, value_vm, value_lsm, abs_diff, vm_units, diff_explanation) |>
  arrange(metric_name)

kable(
  patch_diffs,
  digits = 3,
  caption = "Patch-level metrics with notable differences",
  col.names = c("Metric", "Class", "VM Value", "LSM Value", "Abs Diff", "Units", "Explanation")
)
```

### Class level differences

```{r class-differences}
#| code-fold: true
class_diffs <- class_metrics |>
  filter(abs_diff > 0.001) |>
  arrange(desc(abs_diff)) |>
  group_by(metric_name) |>
  slice(1) |>
  ungroup() |>
  select(metric_name, class, value_vm, value_lsm, abs_diff, vm_units, diff_explanation) |>
  arrange(metric_name)

kable(
  class_diffs,
  digits = 3,
  caption = "Class-level metrics with notable differences",
  col.names = c("Metric", "Class", "VM Value", "LSM Value", "Abs Diff", "Units", "Explanation")
)
```

### Landscape level differences

```{r landscape-differences}
#| code-fold: true
landscape_diffs <- landscape_metrics |>
  filter(abs_diff > 0.001) |>
  select(metric_name, value_vm, value_lsm, abs_diff, vm_units, diff_explanation) |>
  distinct(metric_name, .keep_all = TRUE) |>
  arrange(metric_name)

kable(
  landscape_diffs,
  digits = 3,
  caption = "Landscape-level metrics with notable differences",
  col.names = c("Metric", "VM Value", "LSM Value", "Abs Diff", "Units", "Explanation")
)
```

### Common sources of differences

The main reasons for metric differences across representation types (when they do differ):

1. **Boundary inclusion in edge metrics**: 
   - Vector: Naturally include landscape boundary as part of polygon geometry (default behavior)
   - Raster: Typically exclude boundary edges from calculations
   - Affected metrics: `te`, `ed`, `shape`, `shape_mn`, `shape_sd`, `circle`, `circle_mn`, `circle_sd`, `circle_cv`, `lsi`

2. **Circle vs. square standardization**: 
   - Vector (CIRCLE metrics): Standardize to the smallest circumscribing circle
   - Raster: Standardize to a circumscribing square
   - Affected metrics: `circle`, `circle_mn`, `circle_sd`, `circle_cv`

3. **Core area detection method**:
   - Vector: Contracts patch boundaries inward by a specified distance (in map units)
   - Raster: Excludes edge cells through erosion filtering (in cells)
   - Affected metrics: `ncore`, `cai`, `cai_mn`, `cai_sd`, `cai_cv`, `tca`, `dcore_mn`, `dcore_sd`, `dcore_cv`, `dcad`, `ndca`, `cpland`, `core_cv`

4. **Distance measurement reference points**:
   - Vector: Measures from exact polygon boundaries (continuous geometry)
   - Raster: Measures from cell centers on patch edges (discrete, grid-aligned points)
   - Affected metrics: `enn`, `enn_mn`, `enn_sd`, `enn_cv`

<!-- Since both packages use the same units, residual differences arise from:

- **Geometry representation**: Vector uses exact polygon boundaries; raster approximates with cell edges
- **Patch delineation**: Vector uses polygon boundaries (exact); raster uses cell connectivity (4- or 8-connected)
- **Area measurement**: Vector computes exact polygon area; raster sums cell areas, creating boundary effects
- **Perimeter calculation**: Vector uses exact boundary length; raster counts shared cell edges (creates stair-step approximation)
- **Core area detection**: Vector uses negative buffer; raster uses erosion filter (different geometric operations)
- **Diversity metrics**: Depend on class area proportions, which differ due to rasterization effects on area measurement -->

## vectormetrics-exclusive metrics

**vectormetrics** includes 61 unique metrics not available in **landscapemetrics**. These exclusive metrics reflect the strengths of working with precise vector geometry and include advanced shape complexity measures, specialized distance calculations, and polygon-specific geometry metrics.

```{r exclusive-metrics}
#| code-fold: true
# Load vectormetrics metrics reference
vm_metrics <- vectormetrics::vm_metrics

# metric_map already loaded in setup chunk above
# Extract metrics that exist only in vectormetrics (lsm_fun is NA)
exclusive_metrics <- metric_map |>
  filter(is.na(lsm_fun)) |>
  select(metric_name, level, vm_fun) |>
  distinct() |>
  left_join(vm_metrics |> select(function_name, type, description), by = c("vm_fun" = "function_name")) |>
  select(metric_name, level, type, description) |>
  mutate(level = factor(level, levels = c("patch", "class", "landscape"))) |>
  arrange(metric_name)

kable(
  exclusive_metrics,
  caption = sprintf("All %d vectormetrics-exclusive metrics", nrow(exclusive_metrics)),
  col.names = c("Metric", "Level", "Type", "Description")
)
```

## Conclusion

**vectormetrics** and **landscapemetrics** are complementary tools for landscape ecology:

- **78 shared metrics** provide common ground for landscape analysis, with most showing close agreement when applied to the same landscape in vector and raster formats
- **61 **vectormetrics**-exclusive metrics** offer specialized shape complexity measures suited to vector polygon geometry
- **Systematic differences** (perimeter, edge density, core areas) reflect fundamental differences in how vector and raster representations capture landscape structure

The choice between packages should be driven by your data format and analytical goals -- they represent different but equally valid approaches to landscape metrics. 
